---
title: "Posterior Predictive Checking with predictCheckR"
author: "predictCheckR Maintainer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Posterior Predictive Checking with predictCheckR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 120,
  out.width = "100%"
)
```

# Introduction

Posterior predictive checking (PPC) is a principled, simulation-based
technique for assessing Bayesian model fit.  Rather than relying solely on
likelihood-based information criteria, PPC asks a direct question:

> *If the model is correct, would it generate data that look like the data
> we actually observed?*

## The Posterior Predictive Distribution

Given a Bayesian model with likelihood $p(y \mid \theta)$ and prior
$p(\theta)$, inference yields the posterior $p(\theta \mid y) \propto
p(y \mid \theta)\, p(\theta)$.  The **posterior predictive distribution** for
a future — or hypothetically replicated — outcome $y^{rep}$ is

$$
p(y^{rep} \mid y)
  = \int p(y^{rep} \mid \theta)\, p(\theta \mid y)\, d\theta.
$$

In practice the integral is approximated by Monte Carlo averaging over $S$
posterior draws $\theta^{(1)}, \ldots, \theta^{(S)}$:

$$
p(y^{rep} \mid y)
  \approx \frac{1}{S} \sum_{s=1}^{S} p\!\left(y^{rep} \mid \theta^{(s)}\right).
$$

Each draw $y^{rep(s)}$ is a complete replicated dataset of the same size as
$y$.  The collection $\{y^{rep(s)}\}_{s=1}^{S}$ forms the **posterior
predictive ensemble** that is used for visual and numerical diagnostics.

## Predictive Discrepancy Statistics

A scalar **test statistic** $T(\cdot)$ summarises some property of a dataset
(e.g.\ the mean, standard deviation, or a tail quantile).  The
**Bayesian $p$-value** for $T$ is

$$
p_B = \Pr\!\left(T(y^{rep}) \geq T(y) \mid y\right)
    \approx \frac{1}{S}\sum_{s=1}^{S}
            \mathbf{1}\!\left[T(y^{rep(s)}) \geq T(y)\right].
$$

A value of $p_B \approx 0.5$ indicates that the model is well-calibrated with
respect to $T$.  Extreme values ($p_B < 0.05$ or $p_B > 0.95$) flag potential
misfit.

---

# Package Setup

Install predictCheckR from GitHub (once published) or load it from the source
directory during development:

```{r load-package, message = FALSE}
library(predictCheckR)
library(ggplot2)
```

---

# The Example Dataset

**predictCheckR** ships with `example_data`, a small simulated regression
dataset generated from the model

$$
x_i \sim \mathcal{N}(0, 1), \qquad
y_i = 2 + 3 x_i + \varepsilon_i, \qquad
\varepsilon_i \sim \mathcal{N}(0, 1), \qquad i = 1, \ldots, 100.
$$

```{r load-data}
data(example_data)
head(example_data)
```

```{r plot-data, fig.cap = "Scatter plot of the example regression dataset."}
ggplot(example_data, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30",
              fill = "grey80", linewidth = 0.8) +
  labs(
    title    = "Example Regression Dataset",
    subtitle = "True model: y = 2 + 3x + N(0,1)",
    x        = "Predictor  x",
    y        = "Response  y"
  ) +
  theme_ppc()
```

---

# Generating Fake Posterior Draws

In a real analysis, posterior draws would come from a sampler such as
**rstan** or **brms**.  Here we simulate draws directly to keep the vignette
self-contained.

Suppose we have obtained $S = 400$ posterior samples of the intercept
$\alpha$ and slope $\beta$.  We centre the draws on the true values but add
realistic posterior uncertainty.

```{r fake-draws}
set.seed(101)
S <- 400   # posterior iterations
n <- nrow(example_data)

# Simulate posterior draws: alpha ~ N(2, 0.15^2), beta ~ N(3, 0.12^2)
alpha_draws <- rnorm(S, mean = 2.0, sd = 0.15)
beta_draws  <- rnorm(S, mean = 3.0, sd = 0.12)
sigma_draws <- abs(rnorm(S, mean = 1.0, sd = 0.08))  # residual SD draws

# Posterior draws matrix (S x 2): columns are [alpha, beta]
posterior_draws <- cbind(intercept = alpha_draws, slope = beta_draws)

# Design matrix (n x 2): [1, x]
X <- cbind(1, example_data$x)

dim(posterior_draws)
dim(X)
```

---

# Generating Posterior Predictive Samples

`simulate_ppc()` takes the $S \times P$ draws matrix and an optional design
matrix to produce the $S \times n$ replicated data matrix.

```{r simulate-ppc}
y_rep <- simulate_ppc(
  posterior_draws = posterior_draws,
  X               = X,
  family          = "gaussian",
  sigma_posterior = sigma_draws
)

dim(y_rep)   # 400 rows (draws) x 100 columns (observations)
```

Each row of `y_rep` is one plausible dataset that the model could have
generated.

---

# Running Posterior Predictive Diagnostics

`ppc_diagnostics()` computes several scalar discrepancy statistics and returns
an S3 object with a dedicated `print` method.

```{r diagnostics}
y_obs <- example_data$y

diag_result <- ppc_diagnostics(y_obs = y_obs, y_rep = y_rep)
print(diag_result)
```

**Interpreting the output:**

| Statistic | Interpretation |
|-----------|---------------|
| Mean difference | Should be close to 0. |
| Variance difference | Should be close to 0. |
| Bayesian $p$-value | Values near 0.5 indicate calibration. |
| RMSE | Lower is better; reflects point-prediction accuracy. |
| Coverage (95 % CI) | Should be near 0.95 for a well-calibrated model. |

---

# Visualisation

## Density Overlay

The density overlay is the workhorse visualisation for PPC.  It
superimposes the kernel density estimate of $y^{obs}$ (dark line) over a
random subsample of $y^{rep}$ densities (light lines).

```{r density-overlay, fig.cap = "PPC density overlay. The dark line is the observed data; light lines are 50 random posterior predictive replicates."}
plot_ppc_overlay(
  y_obs     = y_obs,
  y_rep     = y_rep,
  n_samples = 50
)
```

Close agreement between the dark and light lines indicates that the model
reproduces the marginal distribution of the response faithfully.

## Test-Statistic Distribution

`plot_ppc_stat()` shows the distribution of $T(y^{rep})$ across all draws as
a histogram, with a vertical line for $T(y^{obs})$.

```{r ppc-stat-mean, fig.cap = "PPC statistic plot for the mean."}
plot_ppc_stat(y_obs = y_obs, y_rep = y_rep, stat = "mean")
```

```{r ppc-stat-sd, fig.cap = "PPC statistic plot for the standard deviation."}
plot_ppc_stat(y_obs = y_obs, y_rep = y_rep, stat = "sd")
```

The vertical line (observed statistic) should fall well within the
histogram mass when the model is well-specified.

---

# Model Comparison

`compare_models_ppc()` produces a side-by-side table of predictive
performance metrics for two competing models, making it straightforward to
assess which model better describes the data.

```{r model-comparison}
# Mis-specified model: draws centred on a shifted mean
set.seed(202)
posterior_draws_bad <- cbind(
  intercept = rnorm(S, mean = 4.0, sd = 0.15),   # wrong intercept
  slope     = rnorm(S, mean = 3.0, sd = 0.12)
)

y_rep_bad <- simulate_ppc(
  posterior_draws = posterior_draws_bad,
  X               = X,
  family          = "gaussian",
  sigma_posterior = sigma_draws
)

compare_models_ppc(
  y_obs       = y_obs,
  y_rep1      = y_rep,
  y_rep2      = y_rep_bad,
  model_names = c("Correct Model", "Shifted Intercept")
)
```

The correctly-specified model should show lower RMSE and MAE, and a
predictive variance gap close to zero.

---

# Conclusion and Interpretation

Posterior predictive checking provides a transparent, simulation-based window
into model adequacy.  The **predictCheckR** workflow consists of four steps:

1. **Fit** a Bayesian model and extract $S$ posterior draw vectors.
2. **Simulate** posterior predictive replicates with `simulate_ppc()`.
3. **Diagnose** discrepancies with `ppc_diagnostics()`.
4. **Visualise** with `plot_ppc_overlay()` and `plot_ppc_stat()`.

When the model is well-specified, replicated data should be statistically
indistinguishable from the observed data.  Systematic departures — heavy
tails, multimodality, bias in a test statistic — point directly to the
aspects of the data-generating process that the model fails to capture.

---

# Session Information

```{r session-info}
sessionInfo()
```
